Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-21 manuscript_agu_rev1.tex
Fix grammar

\begin{abstract}
Global storm resolving models (GSRMs) represent the next generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but necessarily utilizing microphysics and turbulence parameterizations. Standard-resolution (10–100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using approximately 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010–2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500~m, associated with underestimated lower tropospheric stability and overestimated lifting condensation level. The reanalyses strongly underestimate fog and very low-level clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in MERRA-2, implying a ``too few, too bright'' cloud problem. SO cloud and fog biases are a substantial issue in the analyzed models and result in shortwave and longwave radiation biases.
\end{abstract}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subset our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat–CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat–CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the research vessel (RV) \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the models (ICON, MERRA-2, and ERA5), we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 [ERA5; \citeA{era5}] and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

Some of the observational data were likely used in reanalysis assimilation. The Macquarie Island station surface measurements and radiosonde profiles (not used in our analysis) were sent to the World Meteorological Organization Global Telecommunication System (GTS). The measurements on the RSV \textit{Aurora Australis} and HMNZS \textit{Wellington} were not used outside of research purposes. The AWS measurements, but not lidar or radiosonde measurements on the RV \textit{Tangaroa} voyages, were collected by the New Zealand MetService and communicated to the GTS. The ceilometer measurements on NBP1704 were not used outside of research purposes.

@@

The Vaisala CL51 and CT25K (photos in Fig.~\ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910~nm and 905~nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4~km (CL51), 7.7~km (CL31 emulation mode with 5~m vertical resolution), and 7.5~km (CT25K). The vertical resolution is 10~m (5~m configurable) in CL51 and 30~m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6~s for CL51 on AA15‐16, 16~s for CT25K on MARCUS and MICRE, 36~s for CL51 on RV \emph{Polarstern}, and about 2.37~s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910~nm are both affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910~nm affected more strongly, but we do not expect this to be a significant issue, as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to Network Common Data Form (NetCDF) with cl2nc \cite{kuma2024a} and then processed with the ALCF (Section~\ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to the CL51, equivalent calculations as for the CL51 were done assuming a wavelength of 910~nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.~\ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

@@

The Lufft CHM~15k (photo in Fig.~\ref{fig:map}c) ceilometer operates at a near-infrared wavelength of 1064~nm. The maximum range is 15.4~km; the vertical resolution is 5~m in the near range (up to 150~m) and 15~m above; the sampling (temporal) resolution is 2~s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section~\ref{sec:alcf}) to again produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM~15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

The Automatic Lidar and Ceilometer Framework (ALCF) is a ground-based lidar simulator and a tool for processing observed lidar data, supporting various instruments and models \cite{kuma2021}. It performs radiative transfer calculations to derive equivalent lidar AVBC from an atmospheric model, which can then be compared with observed AVBC. For this purpose, it takes the cloud fraction, liquid and ice mass mixing ratio, temperature, and pressure model fields as an input and is run offline (on the model output rather than inside the model code). The lidar simulator in the ALCF is based on the instrument simulator Cloud Feedback Model Intercomparison Project (CFMIP) Observation Simulator Package (COSP) \cite{bodas-salcedo2011}. After AVBC is calculated, a cloud mask, cloud occurrence by height, and the total cloud fraction are determined. The total cloud fraction is defined as the fraction of profiles with clouds at any height in the lidar cloud mask. The ALCF has been used by several research teams for model and reanalysis evaluation \cite{kuma2020,kremser2021,guyot2022,pei2023,whitehead2023,mcdonald2024a}.

@@

How attenuation of the lidar signal affects cloud detection is dependent on factors such as the optical thickness of the measured cloud and its backscattering phase function, as well as the range-dependent noise standard deviation \cite{kuma2021}. A rough estimate can be made under an assumption of a relatively strongly backscattering cloud of $\beta = \mathrm{100\times 10^{-6} m^{-1}sr^{-1}}$ at a height of $r_1$ = 2 km, range-dependent noise $\beta_n$ at $r_2$ = 8 km of about $\mathrm{5\times 10^{-6} m^{-1}sr^{-1}}$, and cloud detection threshold $\beta_t = \mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$, noise multiplication factor $f$ = 5, at full attenuation the two-way attenuation factor $A$ satisfies $A\beta = \beta_t + f\times \beta_n\left(\frac{r_1}{r_2}\right)^2$, which is equivalent to exponential decay ($A = e^{-2\delta}$) with optical depth $\delta$ (at the lidar wavelength) of about 1.7.

@@

We analyzed a total of about 2400 days of lidar and 2300 radiosonde observations from 31 campaigns and the Macquarie Island subantarctic station, covering the Atlantic, Australian, and New Zealand sectors of the SO over 10 years. This dataset, together with the use of a ground-based lidar simulator, provided a comprehensive basis for evaluating SO cloud and thermodynamic profile biases in the GSRM ICON and the ERA5 and MERRA-2 reanalyses. Our analysis provides a unique evaluation perspective, complementary to satellite observations for evaluating boundary layer clouds and fog, which are predominant in this region. We did not, however, analyze the cloud phase based on ground-based observations. Cloud phase can have a strong impact on the SW radiative transfer due to larger and therefore less numerous hydrometeors (for the same amount of water) scattering much less SW radiation. Especially, the underestimation of fog or very low-level clouds is very substantial in the reanalyses, and we showed that this relates to cloud formation or persistence at RH between 80 and 100\% in the boundary layer in the observations, while in models RH values less than 100\% are associated with clear bins. We subset the dataset by low and high latitude SO bands, cyclonic activity, and stability in order to identify how these conditions influence the biases. The main identified biases are summarized in Table~\ref{tab:summary} and discussed below.

>>>>

\begin{abstract}
Global storm resolving models (GSRMs) represent the next generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but necessarily utilizing microphysics and turbulence parameterizations. Standard-resolution (10–100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using approximately 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a Macquarie Island station during 2010–2021, compared to the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500~m, associated with underestimated lower tropospheric stability and overestimated lifting condensation level. The reanalyses strongly underestimate fog and very low-level clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in MERRA-2, implying a ``too few, too bright'' cloud problem. SO cloud and fog biases are a substantial issue in the analyzed models and result in shortwave and longwave radiation biases.
\end{abstract}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, ICON, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, ICON relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with ICON and reanalyses. We subset our results by cyclonic activity and stability. We found that ICON and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog and very low-level clouds are especially underestimated in the reanalyses. However, ICON and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are operational any more, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat–CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat–CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the research vessel (RV) \emph{Polarstern} and other ships and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and a sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the models (ICON, MERRA-2, and ERA5), we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 [ERA5; \citeA{era5}] and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

Some of the observational data were likely used in the assimilation of the reanalyses. The Macquarie Island station surface measurements and radiosonde profiles (not used in our analysis) were sent to the World Meteorological Organization Global Telecommunication System (GTS). The measurements on the RSV \textit{Aurora Australis} and HMNZS \textit{Wellington} were not used outside of research purposes. The AWS measurements, but not lidar or radiosonde measurements on the RV \textit{Tangaroa} voyages, were collected by the New Zealand MetService and communicated to the GTS. The ceilometer measurements on NBP1704 were not used outside of research purposes.

@@

The Vaisala CL51 and CT25K (photos in Fig.~\ref{fig:map}d, e) are ceilometers operating at near-infrared wavelengths of 910~nm and 905~nm, respectively. The CL51 can also be configured to emulate the Vaisala CL31. The maximum range is 15.4~km (CL51), 7.7~km (CL31 emulation mode with 5~m vertical resolution), and 7.5~km (CT25K). The vertical resolution is 10~m (5~m configurable) in CL51 and 30~m in CT25K observations. The sampling (temporal) resolution is configurable, and in our datasets, it is approximately 6~s for CL51 on AA15‐16, 16~s for CT25K on MARCUS and MICRE, 36~s for CL51 on RV \emph{Polarstern}, and about 2.37~s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905 and 910~nm are both affected by water vapor absorption of about 20\% in the mid-latitudes \cite{wiegner2015,wiegner2019}, with 910~nm affected more strongly, but we do not expect this to be a significant issue, as explained in \citeA{kuma2021}. The instrument data files containing raw uncalibrated backscatter were first converted to the Network Common Data Form (NetCDF) with cl2nc \cite{kuma2024a} and then processed with the ALCF (Section~\ref{sec:alcf}) to produce absolutely calibrated attenuated volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by height, and the total cloud fraction. Because the CT25K uses a very similar wavelength to the CL51, equivalent calculations as for the CL51 were done assuming a wavelength of 910~nm. The Vaisala CL51 and CT25K instruments were used on most of the voyages and stations analyzed here. Fig.~\ref{fig:example}a shows an example of AVBC derived from the CL51 instrument data.

@@

The Lufft CHM~15k (photo in Fig.~\ref{fig:map}c) ceilometer operates at a near-infrared wavelength of 1064~nm. The maximum range is 15.4~km; the vertical resolution is 5~m in the near range (up to 150~m) and 15~m above; the sampling (temporal) resolution is 2~s; and the number of vertical levels is 1024. NetCDF files containing uncalibrated backscatter produced by the instrument were processed with the ALCF (Section~\ref{sec:alcf}) to produce AVBC, cloud mask, cloud occurrence by height, and the total cloud fraction. The CHM~15k was used on four voyages (HMNZSW16, TAN1702, TAN1802, and NBP1704).

@@

The Automatic Lidar and Ceilometer Framework (ALCF) is a ground-based lidar simulator and a tool for processing observed lidar data, supporting various instruments and models \cite{kuma2021}. It performs radiative transfer calculations to derive equivalent lidar AVBC from an atmospheric model, which can then be compared with observed AVBC. For this purpose, it takes the cloud fraction, liquid and ice mass mixing ratio, temperature, and pressure model fields as an input and is run offline (on the model output rather than inside the model code). The lidar simulator in the ALCF is based on the instrument simulator Cloud Feedback Model Intercomparison Project (CFMIP) Observation Simulator Package (COSP) \cite{bodas-salcedo2011}. After AVBC is calculated, a cloud mask, cloud occurrence by height, and the total cloud fraction are determined. The total cloud fraction is defined as the fraction of profiles with clouds at any height in the lidar cloud mask. The ALCF has in the past been used by several research teams for model and reanalysis evaluation \cite{kuma2020,kremser2021,guyot2022,pei2023,whitehead2023,mcdonald2024a}.

@@

How attenuation of the lidar signal affects cloud detection is dependent on factors such as the optical thickness of the measured cloud and its backscattering phase function, as well as the range-dependent noise standard deviation \cite{kuma2021}. A rough estimate can be made under an assumption of a relatively strongly backscattering cloud of $\beta = \mathrm{100\times 10^{-6} m^{-1}sr^{-1}}$ at a height of $r_1$ = 2 km, range-dependent noise $\beta_n$ at $r_2$ = 8 km of about $\mathrm{5\times 10^{-6} m^{-1}sr^{-1}}$, and cloud detection threshold $\beta_t = \mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$, noise multiplication factor $f$ = 5. At full attenuation (relative to the detection threshold), the two-way attenuation factor $A$ satisfies $A\beta = \beta_t + f\times \beta_n\left(\frac{r_1}{r_2}\right)^2$. This is equivalent to exponential decay ($A = e^{-2\delta}$) with optical depth $\delta$ (at the lidar wavelength) of about 1.7.

@@

We analyzed a total of about 2400 days of lidar and 2300 radiosonde observations from 31 campaigns and the Macquarie Island sub-Antarctic station, covering the Atlantic, Australian, and New Zealand sectors of the SO over 10 years. This dataset, together with the use of a ground-based lidar simulator, provided a comprehensive basis for evaluating SO cloud and thermodynamic profile biases in the GSRM ICON and the ERA5 and MERRA-2 reanalyses. Our analysis provides a unique evaluation perspective, complementary to satellite observations for evaluating boundary layer clouds and fog, which are predominant in this region. We did not, however, analyze the cloud phase based on ground-based observations. Cloud phase can have a strong impact on the SW radiative transfer due to larger and therefore less numerous hydrometeors (for the same amount of water) scattering much less SW radiation. Especially, the underestimation of fog or very low-level clouds is very substantial in the reanalyses, and we showed that this relates to cloud formation or persistence at RH between 80 and 100\% in the boundary layer in the observations, while in models RH values less than 100\% are associated with clear bins. We subset the dataset by low and high latitude SO bands, cyclonic activity, and stability in order to identify how these conditions influence the biases. The main identified biases are summarized in Table~\ref{tab:summary} and discussed below.
